### 2018-07-20 Manuscripts/cog-psych/revision/cogpsych.tex
major update. many things. to ken/xian

% ------- ABSTRACT PAGE ------- %
\doublespacing
\section*{Abstract}
\jlanote{need to update} 
\sxlnote{updated} 
%
The ability to creatively generate new concepts and ideas is among the most
fascinating aspects of human cognition, but we do not have a strong
understanding of the cognitive processes and representations underlying concept
generation. In this paper, we study the generation of new categories using the
computational and behavioral toolkit of traditional artificial category
learning. Previous work in this domain has focused on how the statistical
structure of known categories generalizes to generated categories, overlooking
the degree of contrast between the known and generated categories as a factor.
We report three experiments demonstrating that contrast between what is known
and what is created is of fundamental importance for categorization and the
creative process. We propose two novel approaches to modeling category contrast:
one focused on exemplar dissimilarity and another on the representativeness
heuristic. These new models are compared to two key alternatives. Our
experiments and simulations demonstrate specifically how category contrast
influences creative generation, and how it can interact with other constraints
to produce different types of creations. Our work also serves as a concrete
example of how well-established and highly controlled experimental frameworks
(such as those in the field of category learning) can be used to make progress
in an intriguing yet understudied domain.

@@

Previous work in category learning and generation has established that people
are highly sensitive to the structural properties of categories, such as
correlations between the features of category members and the relation between
items within the same category and those in different categories
\citep{roschmervis1975,regier2007,shepard1961learning,smith1995creative}.
Inspired by this work, previous research on the topic of category generation has
explored a similar principle: People tend to create new categories that have
similar {\em statistical regularities} as previously learned categories
\citep{jern2013probabilistic,ward1994structured}. Although this is an important
characteristic of generating new categories, it cannot be the only one. Taken to
the extreme, the best ``new'' category in terms of having the same statistical
regularities to other categories would be identical to a known category that is
representative of the domain (and thus, not new at all).

\sxlnote{I think we can address Charles' point on real world examples of how
  contrast can be important around here.}
To successfully generate something novel, what is generated must be different
from what is already known. This fundamental constraint, ``being different'', or
contrasting from other categories in the relevant domain, is the focus of our
work. Although implicitly assumed in some work, this constraint has been
overlooked in previous research: To our knowledge, there has not been any
systematic investigation addressing how generated categories {\em differ} from
what is already known. Although the idea of category contrast is ubiquitous
throughout the categorization literature, and extends to a variety of other
fields \citep[e.g., color;][]{regier2007}, the idea that a new category should
be ``different'' is vague, as there are many ways it could be different from a
previously observed category.

We examine two different definitions of how categories should differ from one
another: exemplar dissimilarity or representativeness of the alternative
category. To formalize the first technique, we build on the largely successful
exemplar modeling framework that
\citep{medin1978context,nosofsky1984choice,nosofsky1986attention}. In doing so,
we propose a novel exemplar model of category generation, {\em Producing Alike
  and Contrasting Knowledge using Exemplar Representations} (PACKER),
formalizing how new categories should differ from previous categories. It
embodies the exemplar dissimilarity principle by incorporating a factor that
repulses members of opposite categories from one another. To foreshadow, we find
that this notion of contrast predicts human category learning performance, above
and beyond other category learning models in a novel behavioral experiment. This
demonstrates that contrast plays a central role in other important
categorization tasks, which is not specific to category generation.

@@

The outline of the article is as follows. First we describe previous empirical
work on the topic of category generation, as well as the computational
formalizations of the theories in those reports. We then present two hypotheses
for how contrast might affect categorization and formalize them in computational
models. The first is a novel computational model, which is designed to generate
categories that systematically differ from existing categories in the domain.
The second assumes the goal of category generation is to create representative
samples of the opposite of the observed categories. We present two experiments
demonstrating strong and systematic effects of category contrast on concept
generation, and we qualitatively and quantitatively analyze the performance of
each model in capturing human category generation. Afterwards, we find support
for category contrast predicting the ease at which people learn a pair of novel
categories in an empirical experiment. We conclude with a discussion of the
implications of our results for categorization and creative cognition, and
directions for future work.

@@

Much of what we know about concept generation comes from the foundational
literature on creative cognition. In a series of reports, Ward and colleagues
\citep{ward1995s,ward1994structured,marsh1999inadvertent,ward2002role,smith1993constraining}
established that category generation is highly constrained by prior knowledge:
Generated categories tend to consist of features observed in known categories,
and they tend to exhibit the distributional properties as found in known
categories. In a seminal study, \cite{ward1994structured} asked participants to
generate new species of alien animals by drawing and describing members of the
species. People tended to generate species with the same features as on Earth
(e.g., eyes, legs, wings), and possessing the same feature correlations as on
Earth (e.g., feathers co-occur with wings). Likewise, aliens drawn from the same
species tended to share more features with one another compared to members of
opposite species.

\sxlnote{Added thomas98 description here} It is worth pointing out that the
observation of regularities in distributional properties across categories is
not confined to the work of Ward and colleagues. Most notably, \citet{thomas98}
found that when people were asked to generate feature values for exemplars with
a missing feature (specifically circles of different sizes, each with a radial
line of varying orientation), the resulting set of exemplars tended to share the
same feature correlations as exemplars from previously learned categories. [the
transition to the next point isn't great, though]

@@

\cite{jern2013probabilistic} recently showed that concept generation could be
studied in a more controlled manner through the well-developed methods of an
artificial categorization paradigm \citep[see][for a review]{kurtz2015human}. In
Experiments 3 and 4 of their article, participants were exposed to members of
experimenter-defined categories of "crystals" varying in size, hue, and
saturation. Following a training phase during which the experimenter-defined
categories were learned, participants were asked to generate novel categories of
crystals. In a finding mirroring \cite{ward1994structured},
\cite{jern2013probabilistic} found that participants generated categories with
the same distributional properties as the experimenter-defined categories. For
example, after training on categories with a positive correlation between the
size and saturation features (larger sized crystals were more saturated),
participants generated novel categories with the same positive correlation. This
finding is notable, as it demonstrates that category generation can be studied
in a well-known and highly controlled experimental paradigm.

@@

\cite{jern2013probabilistic} additionally tested a ``copy-and-tweak'' model that
broadly resembles the earlier ``Path of Least Resistance'' account. The core
proposal is that participants generate new items by copying stored examples from
memory and tweaking them to generate something new. The copy-and-tweak model
differs from the Path of Least Resistance account in that it notably omits the
hierarchical organization of categories, as well as selectivity in which
features are changed \citep[both of which are factors in the Path of Least
Resistance account;][]{ward2002role}. Instead, their copy-and-tweak model
corresponds to a direct exemplar-similarity approach
\citep[e.g.,][]{nosofsky1984choice,nosofsky1986attention}, generating new items
according to their similarity to known members of the target category. The
copy-and-tweak model provided a poor account of their results, as the
experiments devised by \citeauthor{jern2013probabilistic} were specifically
designed to challenge it. However, its application is notable as a first step
toward understanding theories developed in the creative cognition literature
using well-known formal approaches from the categorization literature.

@@

Prior work on category generation has explored only one factor: distributional
correspondences between learned and generated categories. As a result most of
the computational, theoretical, and empirical efforts have been focused on
explaining those effects. In this paper, we investigate another important
factor: category contrast. To generate a novel concept, individuals must produce
something that is in some capacity {\em different} from what they already know.
By consequence, we propose that contrast should be a primary constraint that
guides how people create new categories: New categories should be different from
existing ones.

\jlanote{check to make sure the next paragraph has been updated enough}
\sxlnote{I think it's updated enough! If given a chance I might edit this a
  little to be less harsh on hierarchical bayesian models in general though I
  think, since the representativeness model is one of them?}
Although it is evident that people are {\em capable} of creating new concepts
and categories, it is not entirely clear how new concepts are systematically
made different from what is already known. The hierarchical Bayesian model
developed by \cite{jern2013probabilistic} assumes that differences between
observed and generated categories are only due to random variation. The model
assumes that generated categories are sampled from the same underlying domain
distribution as observed categories, and will thus share a common distributional
structure. The model does not make predictions about the {\em location} of the
category within the domain (the perceptual instantiation of category members).
Under a strict interpretation of their model, given knowledge of a single
category within the domain, the most probable new category to be generated is
located in {\em exactly} the same location and possesses an identical
distributional structure. This is not an issue with their model specifically,
but with a broader class of standard hierarchical Bayesian models \citep[e.g.,
][]{griffithsscn07,kemp07}. Many of these models assume that at some point of
the latent generative process the same underlying distribution generates all of
the categories and thus, any differences between categories are due to {\em
noise} and should not be {\em systematic}. If the goal of creating a new
category given others in a domain is generating from the posterior predictive
distribution then the best a standard hierarchical Bayesian model without a
notion of contrast built into the model can do at capturing contrast is to
assume that the new category is placed uniformly at random over stimulus space.
But, this defeats the purpose of a hierarchy as it is ignored when determining a
new category location!\footnote{It is plausible that some hierarchical Bayesian
model could be created that generates categories different from each other.
However, this model would not be a standard application or extension of most
pre-existing hierarchical Bayesian models. The generative process would need to
include a component that promotes contrast.} Note that this does not mean
location information and contrast cannot be captured using hierarchical Bayesian
models. In fact, our model of contrast as representativeness is a hierarchical
Bayesian model, but with a different goal than predicting the next category.

@@

\sxlnote{Charles suggested we either remove the section below, or move it to the
  discussion. I've tried finding a good place for it in the discussion but it
  doesn't fit very well, and moving it takes away this nice motivation for
  our further investigation into contrast. I think that keeping it here in the
  introduction might work out, provided Charles finds my argument for why the
  regularity assumption is too strong convincing enough. More details a little
  further down.}
\subsection{Novel Analyses Demonstrating Contrast Effects in Prior Work}

Although existing accounts of category generation broadly overlook the role of
category contrast in determining what is novel versus familiar, it was
implicitly assumed that learners in previous experiments were {\em successful}
in creating new categories. Thus, if contrast plays a robust role in category
genration, its effects should be discernable within the experimental results of
these studies. To provide a test of the influence of category contrast within
existing data, we conducted a novel analysis of Experiment 3 from
\cite{jern2013probabilistic}.\footnote{Although \cite{jern2013probabilistic}
reported four experiments, we focus on Experiment 3 because their first two
experiments tested generation of items into known categories, and their fourth
experiment was identical to Experiment 3 but with a restricted generation
space.}

\sxlnote{Added this bit on regularity below - I'm a little undecided as to leave
  it in the main text or place as a gigantic footnote? It is a bit of a
  digression, and it's really just here to address Charles' point.... If we want
  to keep this in the main text, I'll have to amend the following paragraph
  ``Re-analyzing the results...'' to improve the transition.}
%
It is possible, however, that this data can be explained by a process that does
not involve contrast. Specifically, since participants were trained on exemplars
that share the same hue within a category, it is plausible that they generate
exemplars that all share a common hue. If this is the case, given a $6/8 = .75$
probability that a hue distinct from the experimenter-defined categories is
selected, there is a corresponding $.75$ probability (or about 17 out of 22
participants) that the generated category will have no exemplars that share a
hue with any of the experimenter-defined categories. In addition, that there is
a $2/8 = .25$ probability (about 6 out of 22 participants) that the generated
category will have all six exemplars that share a hue with any of the
experimenter-defined categories. This prediction of regularity across categories
is much closer to the data than the prediction from a Binomial model and does
not require any mechanism of category contrast.

However, upon analysing the consistency of hues in the generated categories, it
is clear that participants do not generally produce categories with entirely
similar hues across all exemplars. Specifically, only .55 of all generated
categories comprised exemplars with a common hue. Consequently, the assumption
of regularity across categories may be inappropriate for this data set, leaving
the contrast-based explanation as the most plausible for our purposes.

@@

PACKER (as well as its name) was in part inspired by earlier work from the
categorization literature. Specifically, \cite{hidaka2011packing} argued that
natural categories ``pack'' the values of features such that different
categories fill the domain space with distance between one another, while
keeping items within the same category close together. Inspired by this idea,
PACKER proposes that generation is constrained by both similarity to members of
the target category (the category in which a stimulus is being generated) as
well as similarity to members of other categories: the most desirable generation
candidates are similar to members of the target category and not similar to
members of contrast categories. This is achieved by aggregating similarity
across known exemplars differently according to class membership. The aggregated
similarity $a(y,x)$ between generation candidate $y$ and stored exemplars $x$ is
given by:

@@

PACKER sets $f(x_j)$ depending on exemplar $x_j$'s category membership:
$f(x_j) = \theta_t$ if $x_j$ is a member of the target category, and
$f(x_j) = -\theta_c$ if $x_j$ is a member of a contrast category.
$\theta_t$ and $\theta_c$ are free parameters
($0 \leq (\theta_t,\theta_c) \leq 1$) controlling the trade-off
between within- and between-category similarity. For example, when
$\theta_t = \theta_c = 0.5$, $f(x_j) = 0.5$ for members of the
target category and $f(x_j) = -0.5$ for members of other categories; thus, the
model is likely to generate items that are similar to members of the target
category but are not similar to members of other categories. In this way,
$\theta_t > 0$ with $\theta_c = 0$ produces exclusive consideration
of target-category members, and $\theta_c > 0$ with $\theta_t = 0$
produces exclusive consideration of contrast-category members. The combination
of $\theta_t$ and $\theta_c$ parameters thus specifies a wide
breadth of possible approaches; by fitting it to a dataset, one can describe the
relative roles of between-category contrast and within-category similarity in
generation. See Figure \ref{fig:packer-examples} for an illustration of how
these parameters control the relative influence of within-category similarity
and contrast to other categories when generating a new exemplar.

@@

It is worth noting that PACKER is only one possible exemplar-based account of
category generation within our proposed framework. That is, PACKER places
specific constraints on the possible values of $f(x_j)$, but other
exemplar-based category generation models with drastically different behavior
can be formalized in this framework by imposing alternative constraints. For
example, as will be discussed in more detail below, PACKER is formally
equivalent to the copy-and-tweak model proposed by \cite{jern2013probabilistic}
when $\theta_c = 0$ and $\theta_t > 0$. Likewise, when
$\theta_t = 0$ and $\theta_c > 0$, PACKER can represent a
contrast-only generation mode, relying exclusively on contrast when generating
new categories. Finally, when $f(x_j) < 0$ for all $x_j$ (regardless of class
membership), a ``pure-packing'' approach is yielded, generating items in
unoccupied areas of the domain. Thus, the proposed framework may be used to
describe a wide variety of qualitatively distinct generation strategies.

It is also worth noting that PACKER represents just one of many possible models
that incorporate contrast in category generation. For example, it may be
possible to extend \cite{jern2013probabilistic}'s hierarchical Bayesian model to
include contrast, but this mechanism would be at odds with the standard
hierarchical Bayesian framework. In contrast, these ideas emerge naturally from
the exemplar view: We have not modified any of the core elements of the GCM in
defining the PACKER model, we simply weight exemplar similarity during
aggregation. Thus, beyond formally specifying the role of contrast in creative
generation, the PACKER account allows us to evaluate the well-understood
principles of the exemplar view within the comparatively understudied field of
conceptual generation. Nonetheless, we will discuss integrating our approaches
in the General Discussion.

@@

In this paper, we report simulations using this copy-and-tweak model. This model
fits within the exemplar-based category generation framework defined above, and
is a continuous-dimension adaptation of the model tested by
\cite{jern2013probabilistic}. By formalizing a model family where PACKER and
copy-and-tweak are different parameterizations of models within the same
framework, the comparison between PACKER and copy-and-tweak provides a test of
the explanatory value of the contrast mechanism based on exemplar dissimilarity:
The account provided by copy-and-tweak will only equal that of PACKER if the
contrast mechanism does not offer an advantage (i.e., if $\theta_c > 0$
significantly improves model fits). Note that the purpose of the article is to
explore and formally analyze the role of contrast in category generation and
thus, we leave extending PACKER to incorporate distributional factors \citep[as
explored by][]{jern2013probabilistic} for future work.

@@

An alternate conceptualization of category contrast is the idea of
representativeness -- exemplars are generated such that they are more
representative of the target category and thus more distinct from and less
similar to other categories. We adopt the definition specified by
\cite{tenenbaum01}, where the representativeness $R$ of an item (or in our case,
an exemplar) $y$ is the relative amount of evidence that is provided by $y$ for
a given hypothesis $h$ compared to all other hypotheses $H'$. More formally,
\begin{equation}
  R(x,h) = \log \dfrac{p(x|h)}{\sum_{h'\in H'}{p(x|h')p(h')}} , 
\label{eq:representativeness}
\end{equation}

@@

This Bayesian formalization of representativeness makes it straightforward to
implement into the existing hieriarchical Bayesian model of category generation
developed by \cite{jern2013probabilistic}. Specifically, both models assume that
exemplars are sampled from a given category distribution ($h$ in
Equation \ref{eq:representativeness}). Each category distribution is a
multivariate normal parameterized by a location vector $\mu_k$ and covariance
matrix $\Sigma_k$ (i.e., $h = \mathcal{N}(\mu_k,\Sigma_k)$). The parameters
$\mu_k$ and $\Sigma_k$ are assumed to be samples from a prior
normal-inverse-Wishart (NIW) distribution parameterized by
$\mu_0,\Sigma_0,\kappa,\nu$. Mathematically,
\begin{align}
  \mu_k, \Sigma_k|\mu_0,\Sigma_0,\kappa,\nu &\sim
                                              \textrm{NIW}(\mu_0,\Sigma_0,\kappa,\nu)
                                              \label{eq:mvnsample}\\
  y|C &\sim \mathcal{N}(\mu_k,\Sigma_k) \label{eq:exemplarsample}
\end{align}

@@

With the assumption of the NIW prior, the location vector $\mu_k$ can be computed
with:
\begin{equation}
  \mu_k = \dfrac
    {\kappa\mu_{0} + n_k \bar{x}_k}
    {\kappa + n_k}
    \label{eq:category_mus}
\end{equation}
where $n_k$ is the number of observed exemplars in category $k$ and $\bar{x}_k$
is the observed category mean. Note that if there are no observed exemplars in
the target category (i.e., if the model is generating a completely novel
category), Equation \ref{eq:category_mus} simplifies to $\mu_k = \mu_0$.

The NIW prior also allows us to infer the category covariance matrix $\Sigma_k$
by computing the following:
\begin{equation}
  \Sigma_k = [\Sigma_D\nu + C_k +
    \dfrac
    {\kappa n_k}
    {\kappa + n_k}
    (\bar{x_k}-\mu_k)(\bar{x_k}-\mu_k)^T
  ] (\nu + n_k)^{-1}
  \label{eq:Sigma_y}
\end{equation}
where $C_k$ is the observed category covariance and $\Sigma_D$ is the domain
covariance matrix from which $\Sigma_k$ samples are obtained. We can
infer $\Sigma_D$ from the observed category covariances $C$ and the prior
covariance $\Sigma_0$. Specifically,
\begin{equation}
    \Sigma_D = \Sigma_0 + \sum_{k}{C_k}
\end{equation}

@@

Despite both the original hierarchical Bayesian model and our representativeness
model sharing identical hierarchical structures, their distinct response
processes can yield very different exemplars. Specifically, the hierarchical
Bayesian model emphasizes the generation of exemplars that maintain
distributional commonalities across categories. In contrast, the
representativeness model focuses on the generation of exemplars that are
representative of the underlying distribution for a target category. Generally,
this representativeness mechanism results in the generation of exemplars that
are less similar (i.e., more different) to contrast categories. We see this
illustrated in Figure \ref{fig:representative-examples}, where the
representativeness model displays a strong preference for generating exemplars
from target category 'B' that are further away from the contrast category 'A'.

>>>>

% ------- ABSTRACT PAGE ------- %
\doublespacing
\section*{Abstract}
The ability to generate new concepts and ideas is among the most
fascinating aspects of human cognition, but we do not have a strong
understanding of the cognitive processes and representations underlying concept
generation. In this paper, we study the generation of new categories using the
computational and behavioral toolkit of traditional artificial category
learning. Previous work in this domain has focused on how the statistical
structure of known categories generalizes to generated categories, overlooking
the degree of contrast between the known and generated categories as a factor.
We report three experiments demonstrating that contrast between what is known
and what is created is of fundamental importance for categorization and innovation. We propose two novel approaches to modeling category contrast:
one focused on exemplar dissimilarity and another on the representativeness
heuristic. These new models are compared to two key alternatives. Our
experiments and simulations demonstrate specifically how category contrast
influences creative generation, and how it can interact with other constraints
to produce different types of creations. Our work also serves as a concrete
example of how well-established and highly controlled experimental frameworks
(such as those in the field of category learning) can be used to make progress
in an intriguing yet understudied domain.

@@

Previous work in category learning and generation has established that people
are highly sensitive to the structural properties of categories, such as
correlations between the features of category members and the relation between
items within the same category and those in different categories
\citep{roschmervis1975,regier2007,shepard1961learning,smith1995creative}.
Inspired by this work, previous research on the topic of category generation has
explored a similar principle: People tend to create new categories that have
similar {\em statistical regularities} as previously learned categories
\citep{jern2013probabilistic,ward1994structured}. Although this is an important
characteristic of generating new categories, it cannot be the only one. Taken to
the extreme, the best ``new'' category in terms of having the same statistical
regularities to other categories would be identical to a known category that is
representative of the domain (and thus, not new at all). Further, contrast has played a critical role studying the spread of new ideas in diverse fields, such as marketing \citep{berger16} and sociology \citep{rogers03}.

To successfully generate something novel, what is generated must be different
from what is already known. This fundamental constraint, ``being different'', or
contrasting from other categories in the relevant domain, is the focus of our
work. Although implicitly assumed in some work, this constraint has been
overlooked in previous research: To our knowledge, there has not been any
systematic investigation addressing how generated categories {\em differ} from
what is already known. Although the idea of category contrast is ubiquitous
throughout the categorization literature, and extends to a variety of other
fields \citep[e.g., color;][]{regier2007}, the idea that a new category should
be ``different'' is vague, as there are many ways it could be different from a
previously observed category.

We examine two different definitions of how categories should differ from one
another: exemplar dissimilarity or representativeness of the alternative
category. To formalize the first technique, we build on the largely successful
exemplar modeling framework that
\citep{medin1978context,nosofsky1984choice,nosofsky1986attention}. In doing so,
we propose a novel exemplar model of category generation, {\em Producing Alike
  and Contrasting Knowledge using Exemplar Representations} (PACKER),
formalizing how new categories should differ from previous categories. It
embodies the exemplar dissimilarity principle by incorporating a factor that
repulses members of opposite categories from one another. To foreshadow, we find
that this notion of contrast predicts human category learning performance, above
and beyond other category learning models in a novel behavioral experiment. This
demonstrates that contrast plays a central role in other important
categorization tasks, and it is not specific to category generation.

@@

The outline of the article is as follows. First we describe previous empirical
work on the topic of category generation, as well as the computational
formalizations of their theoretical explanations. We then present two hypotheses
for how contrast might affect categorization and formalize them in computational
models. The first is a novel exemplar model, which is designed to generate
categories that systematically differ from existing categories in the domain.
The second assumes the goal of category generation is to create representative
samples of the opposite of the observed categories. We present two experiments
demonstrating strong and systematic effects of category contrast on concept
generation, and we qualitatively and quantitatively analyze the performance of
each model in capturing human category generation. Afterwards, we find support
for category contrast predicting the ease at which people learn a pair of 
categories in a novel empirical experiment. We conclude with a discussion of the
implications of our results for categorization and creative cognition, and
directions for future work.

@@

Much of what we know about concept generation and contrast comes from the foundational
literature on creative cognition. In a series of reports, Ward and colleagues
\citep{ward1995s,ward1994structured,marsh1999inadvertent,ward2002role,smith1993constraining}
established that category generation is highly constrained by prior knowledge:
Generated categories tend to consist of features observed in known categories,
and they tend to exhibit the distributional properties found in known
categories. In a seminal study, \cite{ward1994structured} asked participants to
generate new species of alien animals by drawing and describing members of the
species. People tended to generate species with the same features as on Earth
(e.g., eyes, legs, wings), and possessing the same feature correlations as on
Earth (e.g., feathers co-occur with wings). Likewise, aliens drawn from the same
species tended to share more features with one another compared to members of
opposite species.

\sxlnote{Added thomas98 description here} \jlanote{still need the contrast result from her work.} It is worth pointing out that the
observation of regularities in distributional properties across categories is
not confined to generating whole new categories, but it also occures when inferring an unobserved feature within traditional categorization paradigms. Most notably, \citet{thomas98}
found that when people were asked to generate feature values for exemplars with
a missing feature (specifically circles of different sizes, each with a radial
line of varying orientation), the resulting set of exemplars tended to share the
same feature correlations as exemplars from previously learned categories. [the
transition to the next point isn't great, though]

@@

\cite{jern2013probabilistic} recently showed that concept generation could be
studied in a more controlled manner through the well-developed methods of an
artificial categorization paradigm \citep[see][for a review]{kurtz2015human}. In
Experiments 3 and 4 of their article, participants were exposed to members of
experimenter-defined categories of "crystals" varying in size, hue, and
saturation. Following a training phase during which the experimenter-defined
categories were learned, participants were asked to generate novel categories of
crystals. In a finding mirroring \cite{ward1994structured},
\cite{jern2013probabilistic} found that participants generated categories with
the same distributional properties as the experimenter-defined categories. For
example, after learning categories with a positive correlation between the
size and saturation features (larger sized crystals were more saturated),
participants generated novel categories with the same positive correlation. This
finding is notable, as it demonstrates that category generation can be studied
in a well-known and highly controlled experimental paradigm.

@@

\cite{jern2013probabilistic} additionally tested a ``copy-and-tweak'' model that
broadly resembles the earlier ``Path of Least Resistance'' account. The core
proposal is that participants generate new items by copying stored examples from
memory and tweaking them to generate something new. The copy-and-tweak model
differs from the Path of Least Resistance account in that it notably omits the
hierarchical organization of categories, as well as selectivity to which
features are changed \citep[both of which are factors in the Path of Least
Resistance account;][]{ward2002role}. Instead, their copy-and-tweak model
corresponds to a direct exemplar-similarity approach
\citep[e.g.,][]{nosofsky1984choice,nosofsky1986attention}, generating new items
according to their similarity to known members of the target category. The
copy-and-tweak model provided a poor account of their results, as the
experiments devised by \citeauthor{jern2013probabilistic} were specifically
designed to challenge it. Its application is notable as a first step
toward understanding theories developed in the creative cognition literature
using well-known formal approaches from the categorization literature.

@@

Prior work on category generation has explored only one factor: distributional
correspondences between learned and generated categories. As a result most of
the computational, theoretical, and empirical efforts have been focused on
explaining those effects. In this paper, we investigate another important
factor: category contrast. To generate a novel concept, individuals must produce
something that is in some capacity {\em different} from what they already know.
As a consequence, we propose that contrast should be a primary constraint that
underlies people's expectations about how a set of categories are organized: Categories should be different from each other.

Although it is evident that people are {\em capable} of creating new concepts
and categories, it is not entirely clear how new concepts are systematically
made different from what is already known. The hierarchical Bayesian model
developed by \cite{jern2013probabilistic} assumes that differences between
observed and generated categories are only due to random variation. The model
assumes that generated categories are sampled from the same underlying domain
distribution as observed categories, and will thus share a common distributional
structure. The model does not make predictions about the {\em location} of the
category within the domain (the perceptual instantiation of category members).
Under a strict interpretation of their model, given knowledge of a single
category within the domain, the most probable new category to be generated is
located in {\em exactly} the same location and possesses an identical
distributional structure. This is not an issue with their model specifically,
but using a broader class of standard hierarchical Bayesian models without any additional features \citep[e.g.,
][]{griffithsscn07,kemp07}. Many of these models assume that at some point of
the latent generative process the same underlying distribution generates all of
the categories and thus, any differences between categories are due to {\em
noise} and should not be {\em systematic}. If the goal of creating a new
category given others in a domain is generating from the posterior predictive
distribution then the best a standard hierarchical Bayesian model without a
notion of contrast built into the model can do at capturing contrast is to
assume that the new category is placed uniformly at random over stimulus space.
But, this defeats the purpose of a hierarchy as it is ignored when determining a
new category location!\footnote{It is plausible that some hierarchical Bayesian
model could be created that generates categories different from each other.
However, this model would not be a standard application or extension of most
pre-existing hierarchical Bayesian models. The generative process would need to
include a component that promotes contrast.} Note that this does not mean
location information and contrast cannot be captured using hierarchical Bayesian
models. In fact, a model instantiating contrast as representativeness is a hierarchical
Bayesian model. But, it requires an additional factor, which in our case is having a different goal than predicting the next category.

@@

@@

PACKER (as well as its name) was in part inspired by earlier work from the
categorization literature \citep{hikada2011packing,stewart2005}. \citet{hidaka2011packing} argued that
natural categories ``pack'' the values of features such that different
categories fill the domain space with distance between one another, while
keeping items within the same category close together. Inspired by this idea,
PACKER proposes that generation is constrained by both similarity to members of
the target category (the category in which a stimulus is being generated) as
well as similarity to members of other categories: the most desirable generation
candidates are similar to members of the target category and not similar to
members of contrast categories. This is achieved by aggregating similarity
across known exemplars differently according to class membership. The aggregated
similarity $a(y,x)$ between generation candidate $y$ and stored exemplars $x$ is
given by:

@@

\jlanote{I'm a bit confused on the theta constraints. it's clear they need to be non-negative, but you don't usually see both needing to be between 0 and 1 but not sum to some value. is this really what we ended up with? Something feels fishy to me with the math.}
PACKER sets $f(x_j)$ depending on exemplar $x_j$'s category membership:
$f(x_j) = \theta_t$ if $x_j$ is a member of the target category, and
$f(x_j) = -\theta_c$ if $x_j$ is a member of a contrast category.
$\theta_t$ and $\theta_c$ are free parameters
($0 \leq (\theta_t,\theta_c) \leq 1$) controlling the trade-off
between within- and between-category similarity. For example, when
$\theta_t = \theta_c = 0.5$, $f(x_j) = 0.5$ for members of the
target category and $f(x_j) = -0.5$ for members of other categories; thus, the
model is likely to generate items that are similar to members of the target
category but are not similar to members of other categories. In this way,
$\theta_t > 0$ with $\theta_c = 0$ produces exclusive consideration
of target-category members, and $\theta_c > 0$ with $\theta_t = 0$
produces exclusive consideration of contrast-category members. The combination
of $\theta_t$ and $\theta_c$ parameters thus specifies a wide
breadth of possible approaches; by fitting it to a dataset, one can describe the
relative roles of between-category contrast and within-category similarity in
generation. See Figure \ref{fig:packer-examples} for an illustration of how
these parameters control the relative influence of within-category similarity
and contrast to other categories when generating a new exemplar.

@@

It is worth noting that PACKER is only one possible exemplar-based account of
category generation within our proposed framework. That is, PACKER places
specific constraints on the possible values of $f(x_j)$, but other
exemplar-based category generation models with drastically different behavior
can be formalized in this framework by imposing alternative constraints. For
example, as will be discussed in more detail below, PACKER is formally
equivalent to the copy-and-tweak model proposed by \cite{jern2013probabilistic}
when $\theta_c = 0$ and $\theta_t > 0$. Likewise, when
$\theta_t = 0$ and $\theta_c > 0$, PACKER can represent a
contrast-only generation mode, relying exclusively on contrast when generating
new categories. When $f(x_j) < 0$ for all $x_j$ (regardless of class
membership), a ``pure-packing'' approach is yielded, generating items in
unoccupied areas of the domain. Thus, the proposed framework may be used to
describe a wide variety of qualitatively distinct generation strategies.

@@

In this paper, we report simulations using this copy-and-tweak model. This model
fits within the exemplar-based category generation framework defined above, and
is a continuous-dimension adaptation of the model tested by
\cite{jern2013probabilistic}. By formalizing a model family where PACKER and
copy-and-tweak are different parameterizations of models within the same
framework, the comparison between PACKER and copy-and-tweak provides a test of
the explanatory value of the contrast mechanism based on exemplar dissimilarity:
The account provided by copy-and-tweak will only equal that of PACKER if the
contrast mechanism does not offer an advantage (i.e., if $\theta_c > 0$
significantly improves model fits). Note that the purpose of the article is to
explore and formally analyze the role of contrast in categorization and
thus, we leave extending PACKER to incorporate distributional factors \citep[as
explored by][]{jern2013probabilistic} for future work.

@@

An alternate conceptualization of category contrast is the idea of
representativeness -- exemplars are generated such that they resemble the target category and thus more distinct from and less
similar to other categories. We adopt the formalization given by
\cite{tenenbaum01}, where the representativeness $R$ of an item (or in our case,
an exemplar) $y$ is the relative amount of evidence that is provided by $y$ for
a given hypothesis $h$ in a space of hypotheses $\mathcal{H}$ in contrast to all other hypotheses $\mathcal{H}^c=\mathcal{H} \setminus \{h\}$:
\begin{equation}
  R(x,h) = \log \dfrac{p(x|h)}{1-p(x|\mathcal{H}^c)} = \log \dfrac{p(x|h)}{\sum_{h'\in H'}{p(x|h')p(h')}} . 
\label{eq:representativeness}
\end{equation}

@@

The Bayesian formalization of representativeness makes it straightforward to
extend the existing hierarchical Bayesian model of category generation
developed by \cite{jern2013probabilistic}.\footnote{We thank Charles Kemp for the suggestion.} Specifically, both models assume that
exemplars are sampled from a given category distribution ($h$ in
Equation \ref{eq:representativeness}). Each category distribution is a
multivariate normal parameterized by a location vector $\mu_k$ and covariance
matrix $\Sigma_k$ (i.e., $h = \mathcal{N}(\mu_k,\Sigma_k)$). The parameters
$\mu_k$ and $\Sigma_k$ are assumed to be samples from a prior
normal-inverse-Wishart (NIW) distribution parameterized by
$\mu_0,\Sigma_0,\kappa,\nu$. Mathematically,
\begin{align}
  \mu_k, \Sigma_k|\mu_0,\Sigma_0,\kappa,\nu &\sim
                                              \textrm{NIW}(\mu_0,\Sigma_0,\kappa,\nu)
                                              \label{eq:mvnsample}\\
  y|C &\sim \mathcal{N}(\mu_k,\Sigma_k) \label{eq:exemplarsample}
\end{align}

@@

With the assumption of the NIW prior, the expected location vector $\mu_k$ given exemplars observed from category $k$ is
\begin{equation}
  \mu_k = \dfrac
    {\kappa\mu_{0} + n_k \bar{x}_k}
    {\kappa + n_k}
    \label{eq:category_mus}
\end{equation}
where $n_k$ is the number of observed exemplars in category $k$ and $\bar{x}_k$
is the observed category mean. Note that if there are no observed exemplars in
the target category (i.e., if the model is generating a completely novel
category), Equation \ref{eq:category_mus} simplifies to $\mu_k = \mu_0$.

The NIW prior also allows us to infer the category covariance matrix $\Sigma_k$
by computing the following:
\begin{equation}
  \Sigma_k = [\nu \Sigma_D+ C_k +
    \dfrac
    {\kappa n_k}
    {\kappa + n_k}
    (\bar{x_k}-\mu_k)(\bar{x_k}-\mu_k)^T
  ] (\nu + n_k)^{-1}
  \label{eq:Sigma_y}
\end{equation}
where $C_k$ is the observed category covariance and $\Sigma_D$ is the domain
covariance matrix from which $\Sigma_k$ samples are obtained. We can
infer $\Sigma_D$ from the observed category covariances $C$ and the prior
covariance $\Sigma_0$. Specifically,
\begin{equation}
    \Sigma_D = \Sigma_0 + \sum_{k}{C_k}
\end{equation}

@@

Despite both the original hierarchical Bayesian model and our representativeness
model sharing identical hierarchical structures, their distinct response
processes can yield very different exemplars. Specifically, the hierarchical
Bayesian model emphasizes the generation of exemplars that maintain
distributional commonalities across categories. In contrast, the
representativeness model focuses on the generation of exemplars that are
representative of the underlying distribution for a target category. \jlanote{The next sentence is weird sounding and should be cleaned up} Generally,
this representativeness mechanism results in the generation of exemplars that
are less similar (i.e., more different) to contrast categories. We see this
illustrated in Figure \ref{fig:representative-examples}, where the
representativeness model displays a strong preference for generating exemplars
from target category 'B' that are further away from the contrast category 'A'.
